NumPy---Array Programming in Python

### 2020-01-27 body.tex
Revise ecosystem

\fixme{Up to here we were talking about arrays, now we introduce array programming.
Need to better explain vectorization in array programming.
This is perhaps the essential thing about it so we should make it slightly
more explicit.  It is just shorter to type, but it is easier to think
about. }
To complement the array syntax, NumPy includes functions that perform
vectorized calculations on arrays, including arithmetic, statistics, and
trigonometry.
They aim to loop over array elements near-optimally, taking into consideration,
e.g., strides, in order to best utilize the computer's fast cache memory.
Vectorized operations that would take many tens of lines to express in
languages such as C can often be implemented as a single, clear Python
expression.

When performing a vectorized operation (such as addition) on two arrays, one
expects that these arrays should have the same shape.
% might expect if you are coming from other languages
% wouldn't expect if you were doing math
% scalar vector multiplication, scalar matrix multiplication, etc.
However, through \emph{broadcasting}, NumPy allows the dimensions to differ, while
still producing results that appeal to intuition.
A trivial example is the addition of a scalar value to an array, but it also
generalizes to more complex examples such as scaling each column of an array,
or generating a grid of coordinates.
In broadcasting, one or both arrays are virtually duplicated (that is, without
copying any data in memory), so that the shapes of the operands match
(Fig.~\ref{fig:array-concepts}d).
Broadcasting is also applied whenever an array is indexed using arrays of
indices (Fig.~\ref{fig:array-concepts}c).

@@

The brevity and expressive power of NumPy's array programming syntax combined with 
an advanced interactive environment like IPython \cite{perez2007ipython}
or Jupyter \cite{Kluyver:2016aa} makes interactive data exploration fluid and simple.
SciPy and Matplotlib are tightly integrated with NumPy to expand the suite of
standard algorithms and provide basic plotting capabilities.
SciPy provides many fundamental algorithms for scientific computing on arrays
including mathematical, scientific, and engineering routines.
Matplotlib generates publication-ready figures and visualizations.

% say something about this already being a good foundation comparable to
% other scientific computing languages.

The scientific Python ecosystem (Fig.~\ref{fig:ecosystem}) builds on top of
this foundation providing many, widely used technique-specific
libraries\cite{pedregosa2011scikit,vanderwalt2014scikit,SciPyProceedings_11}
and several domain-specific projects and libraries
\cite{astropy:2013,astropy:2018,cock2009biopython,millman2007analysis,2018EGUGA..2012146H}.
Exposing these simple array programming primitives, as well as the surrounding ecosystem of
tools, in Python---an interpreted, general purpose language---creates an ideal environment for
interactive, exploratory data analysis where users may iteratively inspect,
manipulate, and visualize their data.
Working interactively with their data, users discover through experimentation
bits of code that they stitch together into imperative or functional programs.
This rich and productive environment has made the scientific Python ecosystem popular for scientific
research.

@@

% need better transition
% similar tools exist in physics, chemistry, biology, earth sciences,
% engineering
% this means there is an extraordinary need for stability and reliability
% this conflicts with competing need to innovate
% resolving this tension requires balancing trade-offs
% this lead to a proliferation of arrays

The rich ecosystem of tools building on NumPy and the larger community
of scientists and programmers using those tools mean that NumPy has a
responsibility to maintain considerable stability.
This need creates interesting challenges for NumPy development as the
computational environment scientists are analyzing their data on continues
to evolve and change.

% - review supporting the ecosystem (actual support element)
%   - docstrings / testing / packaging
%   - coordinating mechanism
%     - deprecation policy

@@

%\fixme{Add references somewhere to Zarr, https://arrow.apache.org/,
%https://parquet.apache.org/,
%https://github.com/ruby-numo/numo-narray/wiki/Comparison-with-Numpy}

% pandas, xarray, ...

@@

The community's efforts to fill the resultant gaps therefore led to a
proliferation of arrays including distributed arrays---which are stored on
multiple computers, GPU arrays---which utilize graphics processing units for
fast computation, computation graphs---which postpone and combine calculations
before executing them efficiently, and sparse arrays---which typically contain
few non-zero values, and store only those in memory.
%distributed arrays, GPU arrays, delayed execution arrays, sparse arrays,
%foreign language arrays, and more.

@@

Primarily, there
exist two types of Python array objects: (1) arrays that mimic NumPy arrays but are
fundamentally different and (2) subclasses of NumPy arrays.

@@

This culture of using best practices for producing reliable scientific software
has been eagerly adopted by the ecosystem of libraries that build on NumPy.
For example, in a recent award given by the Royal Astronomical Society to
Astropy, they state:
\begin{quotation}
\noindent\emph{The Astropy Project has provided hundreds of junior scientists
with experience in professional-standard software development practices
including use of version control, unit testing, code review and issue tracking
procedures. This is a vital skill set for modern researchers that is often
missing from formal university education in physics or astronomy.}
\end{quotation}
Community members explicitly work to address this lack of formal education
through formal courses and workshops
\cite{wilson-software-carpentry,hannay-scientific-software-survey,millman2018teaching}.

>>>>

To complement the array syntax, NumPy includes functions that perform
\emph{vectorized} calculations on arrays, including arithmetic, statistics, and
trigonometry.
Vectorization---operating on whole arrays rather than their individual
elements---is essential to array programming.
This means that operations that would take many tens of lines to express in
languages such as C can often be implemented as a single, clear Python
expression.
This results in concise code and frees users to focus on the details of
their analysis, while NumPy handles looping over array elements near-optimally,
taking into consideration, for example, strides, in order to best utilize the
computer's fast cache memory.

When performing a vectorized operation (such as addition) on two arrays with
the same shape, it is clear what should happen.
Through \emph{broadcasting}, NumPy allows the dimensions to differ, while
still producing results that appeal to intuition.
A trivial example is the addition of a scalar value to an array, but it also
generalizes to more complex examples such as scaling each column of an array,
or generating a grid of coordinates.
In broadcasting, one or both arrays are virtually duplicated (that is, without
copying any data in memory), so that the shapes of the operands match
(Fig.~\ref{fig:array-concepts}d).
Broadcasting is also applied when an array is indexed using arrays of
indices (Fig.~\ref{fig:array-concepts}c).

@@

SciPy and Matplotlib are tightly integrated with NumPy.
% to expand the suite of standard algorithms and
% provide basic plotting capabilities.
SciPy provides fundamental algorithms for scientific computing
including mathematical, scientific, and engineering routines.
Matplotlib generates publication-ready figures and visualizations.
The combination of NumPy, SciPy, and Matplotlib with 
an advanced interactive environment like IPython \cite{perez2007ipython}
or Jupyter \cite{Kluyver:2016aa}
provides a complete foundation for array-based scientific computing
and makes interactive data exploration fluid and simple.
The scientific Python ecosystem (Fig.~\ref{fig:ecosystem}) builds on top of
this foundation providing many, widely used technique-specific
libraries\cite{pedregosa2011scikit,vanderwalt2014scikit,SciPyProceedings_11}
and several domain-specific projects and libraries
\cite{astropy:2013,astropy:2018,cock2009biopython,millman2007analysis,2018EGUGA..2012146H}.
NumPy plays a coordinating role for the ecosystem of array-aware
libraries including setting documentation standards,
providing array testing infrastracture,
and coordinating minimum version depedencies.

Exposing this foundation for array programming, as well as the surrounding ecosystem of
tools, in IPython or Jupyter, creates an ideal environment for
interactive, exploratory data analysis where users iteratively inspect,
manipulate, and visualize their data.
Working interactively with their data, users discover through experimentation
bits of code that they stitch together into imperative or functional programs
or narrative computational notebooks.
This rich and productive environment has made the scientific Python ecosystem popular for scientific
research.
As a result, important scientific software packages build on the ecosystem
further exanding it for specific research projects.

@@

Similar examples can be found in physics, earth sciences, biology,
and more.
These researchers depend on NumPy being correct, reliable,
and repeatable.
The rich ecosystem of tools building on NumPy and the larger community
of scientists and programmers using those tools mean that NumPy has a
responsibility to maintain considerable stability.
This conflicts with the need to innovate;
resolving it requires balancing trade-offs.
% our focus is ...
This need creates interesting challenges for NumPy development as the
computational environment scientists are analyzing their data on continues
to evolve and change.
% this leads to a proliferation of arrays

@@

% pandas, xarray, ...
% distributed arrays, delayed execution arrays
% GPU arrays, tensor computation libraries
% sparse arrays
% foreign language arrays

@@

The community's efforts to fill the resultant gaps therefore led to a
proliferation of arrays including distributed arrays---which are stored on
multiple computers, GPU arrays---which utilize graphics processing units for
fast computation, computation graphs---which postpone and combine calculations
before executing them efficiently, and sparse arrays---which typically contain
few non-zero values, and store only those in memory.

@@

Primarily, there
exist two types of Python array objects: (a) arrays that mimic NumPy arrays but are
fundamentally different and (b) subclasses of NumPy arrays.

@@

This culture of using best practices for producing reliable scientific software
has been adopted by the ecosystem of libraries that build on NumPy.
For example, in a recent award given by the Royal Astronomical Society to
Astropy, they state:
\begin{quotation}
\noindent\emph{The Astropy Project has provided hundreds of junior scientists
with experience in professional-standard software development practices
including use of version control, unit testing, code review and issue tracking
procedures. This is a vital skill set for modern researchers that is often
missing from formal university education in physics or astronomy.}
\end{quotation}
Community members explicitly work to address this lack of formal education
through formal courses and workshops
\cite{wilson-software-carpentry,hannay-scientific-software-survey,millman2018teaching}.
